\abstractEN {
    This thesis deals with the design and implementation of a stacking window manager for the X Window System.
    A window manager is the core component of any modern graphical desktop -- it is responsible
    for the placement and appearance of application windows on the screen.
    While there is a plethora of window managers, especially on the X Window System,
    most of them are either heavyweight window managers that are part of a desktop environment or lightweight tiling managers.
    In this thesis, we try to fill the gap by developing a lightweight stackable window manager that complies with the freedesktop
    standards such as ICCCM and EWMH, and that follows the {\sc unix} philosophy of doing one thing and doing it well.
    The focus has been on simplicity, code readability, testability, and making it easy to use and extend.
}

\abstractCZ {
    Tato práce se zabývá návrhem a implementací klasického (nedlaždicového) správce oken pro X Window System.
    Správce oken se stará o rozmístění a vzhled oken jednotlivých aplikací na obrazovce a je tedy klíčovou součástí
    jakéhokoliv moderního počítače s grafickým rozhraním.
    Ačkoliv již existuje mnoho správců oken, zejména pro X Window System, chybí zde jednoduchý nedlaždicový.
    Tato práce se snaží jeden takový vytvořit.
    Cílem bylo, aby splňoval freedesktop standardy ICCCM a EWMH
    a držel se {\sc unix}ové filozofie dělat jednu věc a dělat ji dobře.
    Důraz byl také kladen na jednoduchost, čitelný a testovatelný kód, a na to,
    aby byl správce oken snadno použitelný a rozšiřitelný.
}

\keywordsEN {
    window manager, stacking window manager, X Window System, X11, Xorg, ICCCM, EWMH, Go, golang, open source
}
