\label[chap:conclusions]
\chap Conclusion

In this thesis, we described the working mechanisms of the X Window System Protocol
and two standard protocols defined on top of it, ICCCM and EWMH.
Those mechanisms and policies were described especially concerning window managers,
special X Server clients that control the placement and appearance of windows.
We also described what a window manager is, its different types,
and features it usually offers, demonstrated on some existing implementations.
Finally, a new stacking window manager called swm was designed in such a way
that it is small, extensible, and scriptable.
It was implemented in Go, a modern system programming language,
is ICCCM and EWMH compliant, and comes with stacking window management,
advanced window grouping mechanism, basic window decorations, and the command-line
utility that makes it easy to manipulate the windows in a simple but powerful way.
We also described how a window manager could be tested and how tests for core features of swm were implemented.

\nonum
\sec Future Work

Swm is an open-source project, its source code is published on GitHub\urlnote{https://github.com/janbina/swm} under the MIT license.
It was designed and implemented in a way to be built upon, to be extended with new functionalities, and adapted to various uses and user preferences.
Therefore, there is a potential for it to be extended by third parties, either in a way of contributions or by starting a whole new
project using swm as a starting point.

%In this thesis, we described the working mechanisms of X Window System Protocol
%and two standard protocols defined on top of it, ICCCM and EWMH.
%%especially in regard to window managers, special X Server clients
%%that control the placement and appearance of windows.
%Those mechanisms and policies were described especially with respect to window managers,
%special X Server clients that control the placement and appearance of windows.
%% VARIANTS ^^^^
%We also described what a window manager is, its different types, and features it usually offers,
%demonstrated on some existing %window manager
%implementations.
%Finally, a new stacking window manager called swm was designed in such a way
%that it is small, extensible, and scriptable.
%It was implemented in Go, a modern system programming language,
%is ICCCM and EWMH compliant, and comes with stacking window management,
%advanced window grouping mechanism, basic window decorations, and its own
%command-line utility that makes it easy to manipulate the windows
%in a simple but powerful way.
%We also described how a window manager could be tested and how tests for core
%features of swm were implemented.

% ???
%Some additions that come to mind may include better window decorations or some mechanism to assign different properties
%to windows upon startup based on application -- for example, web browser always starts maximized.
